---
title: "STAT 408 Final Project"
author: "Aaron Myrold, Sathvik Maridasana Nagaraj, Rolando Santos"
date: "2023-11-26"
output: 
  html_document:
    toc: yes
    theme: default
  pdf_document: default
urlcolor: cyan
---

# Bike Sharing Predictive Model

## Introduction
TBD

### Source
Our data was found from https://archive.ics.uci.edu/dataset/275/bike+sharing+dataset. This data provides 2 files for data, we will be using the 'day.csv' since every entry is an individual day compared to every hour in the 'hour.csv'.

### Variables
The dataset has 16 unique variables, however, this study will start off with using 12 of them.

#### Used
- **season**: Season
  + `1`: Winter
  + `2`: Spring
  + `3`: Summer
  + `4`: Fall
- **year**: Year
  + `0`: 2011
  + `1`: 2012
- **mnth**: Month (`1`: January - `12`: December)
- **holiday**: Whether day is holiday or not (`0`: No, `1`: Yes)
- **weekday**: Day of the Week (`0`: Sunday - `6`: Saturday)
- **workingday**: If day is neither weekend nor holiday (`0`: No, `1`: Yes)
- **weathersit**: Weather Situation
  + `1`: Clear
  + `2`: Misty
  + `3`: Snow/Rain
- **temp**: Normalized Temperature in Celsius
- **atemp**: Normalized Feeling Temperature in Celcius
- **windspeed**: Normalized Wind Speed
- **cnt**: Count of total rental bikes including both casual and registered (Response)

#### Unused
- **instant**: Record Index
- **dteday**: Date
- **casual**: Count of Casual Users
- **registered**: Count of Registered Users

The reason why `casual` and `registered` are not used is because they are too closely related to the response `cnt` due to `cnt` being the sum of both variables and can cause issues with our model due to the extremely high correlation.

Each group member will try to fit several models with the price as the response, and a combination of the remaining variables as predictors. The best model fit by each member will be listed below, and will be discussed in the `Results` and `Discussion` sections that follow.

## Methods

```{r echo = FALSE, message = FALSE, warning = FALSE}
# Libraries Used
library(tidyverse)
library(dplyr)
library(forcats)
library(ggplot2)
library(gridExtra)
library(cowplot)
library(carData)
library(corrgram)
```

```{r}
# Set seed to ensure you always have same random numbers generated
set.seed(408)  
```

```{r}
# Importing the data
bike_sharing <- read.csv("day.csv")
head(bike_sharing)
```

```{r}
na_count <-sapply(bike_sharing, function(y) sum(is.na(y))) 
na_count <- data.frame(na_count)
na_count
```

There are no missing values in the the dataset.

```{r}
bike_sharing <- bike_sharing[ ,!names(bike_sharing) %in% c("instant", "registered", "dteday", "casual")]
head(bike_sharing)
```

Removed the `instant`, `registered`, `dteday` and `casual` columns from our dataset.

### Helper Functions

```{r}
# Plots "Fitted vs Residual" and "Q-Q" Plots 
plot_diagnostics = function(model) {
  par(mfrow = c(1, 2))
  plot(fitted(model), resid(model), 
       col = "grey", pch = 20, 
       xlab = "Fitted", ylab = "Residuals", 
       main = "Data from Additive Model")
  abline(h = 0, col = "darkorange", lwd = 2)
  qqnorm(resid(model), main = "Normal Q-Q Plot", col = "darkgrey")
  qqline(resid(model), col = "dodgerblue", lwd = 2)
}

# Create bar plots for categorical vs numeric relationships
create_bp <- function(var){
  bike_sharing %>%
    ggplot(
      aes_string(x = toString(var), y = "cnt", fill = "year")
    ) + geom_col(
      position = position_dodge()
    )
}

# Calculates RMSE given "actual" and "predicted values"
get_train_rmse = function(model) {
  predicted = (predict(model, bike_sharing_trn))
  sqrt(mean((predicted - bike_sharing_trn$cnt)^2))
}

get_test_rmse = function(model) {
  predicted = (predict(model, bike_sharing_tst))
  sqrt(mean((predicted - bike_sharing_tst$cnt)^2))
}
```

### Investigation and Variable Relationships

First we will map our categorical variables to their respective descriptions for clarity.

```{r}
bike_sharing <- bike_sharing %>% 
  rename(
    year = yr,
    month = mnth,
  )

bike_sharing <- bike_sharing %>% mutate(
  holiday = fct_recode(as.factor(holiday),
    "No" = "0",
    "Yes" = "1",
  )
) %>% mutate(
  season = fct_recode(as.factor(season),
    "Spring" = "1",
    "Summer" = "2",
    "Fall" = "3",
    "Winter" = "4"
  )
) %>% mutate(
  year = fct_recode(as.factor(year),
    "2011" = "0",
    "2012" = "1"
  )
) %>% mutate(
  month = fct_recode(as.factor(month),
    "January" = "1",
    "February" = "2",
    "March" = "3",
    "April" = "4",
    "May" = "5",
    "June" = "6",
    "July" = "7",
    "August" = "8",
    "September" = "9",
    "October" = "10",
    "November" = "11",
    "December" = "12"
  )
) %>% mutate(
  weekday = fct_recode(as.factor(weekday),
    "Sunday" = "0",
    "Monday" = "1",
    "Tuesday" = "2",
    "Wednesday" = "3",
    "Thursday" = "4",
    "Friday" = "5",
    "Saturday" = "6"
  )
) %>% mutate(
  workingday = fct_recode(as.factor(workingday),
    "No" = "0",
    "Yes" = "1"
  )
) %>% mutate(
  weathersit = fct_recode(as.factor(weathersit),
    "Clear" = "1",
    "Misty" = "2",
    "Snow/Rain" = "3"
  )
)
```

Some plots showing the relationship between the categorical variables and `cnt`.

```{r}
plot_grid(
  create_bp("season"), 
  create_bp("weathersit"),
  create_bp("holiday"),
  create_bp("month"),
  labels=c("A", "B"),
  ncol = 2, nrow = 2)
```

Some plots showing the relationship between the numeric variables and `cnt`.

```{r}
#TODO
```

We will also fit a base model to view the current significance of coefficients and to view any constant variance or normality issues.

```{r}
model <- lm(cnt ~ ., data = bike_sharing)
summary(model)
```
One glaring issue with our current base model is that we have a collinearity issue related to `workingday`. Reviewing our variable definitions, `workingday` is directly correlated to whether or not a the current day is neither a weekend or holiday. Thus it would be best to also remove this variable from our model going forward.

```{r}
plot_diagnostics(model)
```

We can see that our data does seem to uphold constant variance, however there appears to be some tail skewedness on our Q-Q plot. Our updated models may resolve this issue.

### Predictive Model Preparation

Since one of the goals of our project is the predict the number of bikes needed for a given day based on provided conditions, it would be best to adjust the categorical variables by turning them into dummy variables/one-hot encoding them.

```{r}
# Creating Dummy Variables
bike_sharing <- fastDummies::dummy_cols(bike_sharing, remove_first_dummy = TRUE)

# Removing the original columns
bike_sharing <- bike_sharing[ ,!names(bike_sharing) %in% c("season", "year", "month", "holiday", "weekday", "workingday", "weathersit")]
head(bike_sharing)
```


We then split the resulting dataset into training and testing datasets, using an 80:20 split.

```{r}
sample_size = floor(0.8 * nrow(bike_sharing)) 
indices = sample(seq_len(nrow(bike_sharing)), size = sample_size)
bike_sharing_trn = bike_sharing[indices,]
bike_sharing_tst = bike_sharing[-indices,]
```

### Models

TBD

## Results

TBD

### Plot Diagnostics

TBD

### MSE & $R^2$ Testing

TBD

## Discussion

TBD

### Plot Diagnostics

TBD

### MSE & $R^2$ Testing

TBD

## Conclusion

TBD